---
title: "GTEx data"
author: "Britta Velten"
date: "30/8/2017"
output:
  BiocStyle::html_document:
    toc: true
---


# Introduction
Script to obtain data from GTEx for analysis of the aging transcriptome.

```{r, message=FALSE, warning=FALSE}
library(data.table)
library('recount')
library(DESeq2)
library(ggplot2)
library(dplyr)
library(magrittr)
library(plyr)
library(reshape2)
options(stringsAsFactors = FALSE)
```

# Data import
```{r}
# adpat to the directory containing the txt files for donor and sample info
datadir <- "~/Documents/LassoVariants/grpRR_model/Applications/GTEx/data/"
```

## Get count data from recount
GTEx data is available on [recount](https://jhubiostatistics.shinyapps.io/recount/).
```{r, eval=FALSE}
url <- download_study('SRP012682')
load(file.path('SRP012682', 'rse_gene.Rdata'))
rse_gene
# save(rse_gene, file=file.path(datadir, "rse_gene.RData"))
```

```{r, eval= FALSE, echo = FALSE}
load(file.path(datadir, "rse_gene.RData"))
```


## Get sample and donor info
This is not publicly available from recount and requires dbGAP access.
```{r, eval= FALSE}
# load files phs000424.v6.pht002742.v6.p1.c1.GTEx_Subject_Phenotypes.GRU.txt and SraRunTable.txt
files <- list.files(datadir)
files <- files[grepl(".txt", files)]
subjectPheno <- fread(file.path(datadir, files[1]), header = TRUE, skip = 10)
sraRun <- fread(file.path(datadir, files[2]), header = TRUE)
```

## Map to colData of RSE object
Sample and donor information is spread across three object: 

- rse_gene 
    * sample SRS389722 (match with sraRun.SRA_Sample_s)
    * experiment SRX222703 (match with sraRun.Experiment_s)
    * run SRR660824 (match with sraRun.Run_s)

- sraRun (10474 runs)
    * Experiment_s SRX197226 (match with rse_gene.experiment)
    * Run_s  SRR595926 (match with rse_gene.run)
    * SRA_Sample_s SRS332984 (match with rse_gene.sample)
    * submitted_subject_id_s GTEX-N7MS (match with subjectPheno.SUBJID)

- subjectPheno (572)
    * SUBJID GTEX-N7M   (match with sraRun.submitted_subject_id_s)

The following code fills in information in colData of rse_gene on donors and samples contained in sraRun and subjectPheno.
```{r, eval=FALSE}
colnames(sraRun)[match(c("Experiment_s", "Run_s", "SRA_Sample_s", "submitted_subject_id_s"),colnames(sraRun))] <- c("experiment", "run", "sample", "subject_id")
colnames(subjectPheno)[match(c("SUBJID"), colnames(subjectPheno))] <- c("subject_id")

allInfo <- join(as.data.frame(colData(rse_gene)), sraRun, by=c("experiment", "run", "sample"), type="left")
allInfo <- join(allInfo, subjectPheno, by=c("subject_id"), type="left")
all(rownames(colData(rse_gene))==allInfo$run)

rse_gene_ext <- rse_gene
colData(rse_gene_ext) <- DataFrame(allInfo)
```


## Save joint data object
```{r, eval=FALSE}
save(rse_gene_ext, file=file.path(datadir, "rse_gene_ext.RData"))
```

```{r}
load(file.path(datadir, "rse_gene_ext.RData"))
```

# Donor - level information
```{r}
cdat <- as.data.frame(colData(rse_gene_ext))
nrow(cdat)
```

Summary on donor level
```{r}
donor_meta <- cdat %>%
  group_by(subject_id) %>%
  dplyr::summarise(n_samples= length(unique(sample)),
                   n_runs= length(unique(run)),
                   n_experiments= length(unique(experiment)),
                   age= unique(AGE),
                   sex=unique(sex_s),
                   n_tissues_d= length(unique(smtsd)),
                   n_tissues= length(unique(smts)),
                   race = unique(RACE),
                   ethnicity = unique(ETHNCTY),
                   height= unique(HGHT),
                   weight= unique(WGHT),
                   BMI = unique(BMI))
nrow(donor_meta)
```


# Overview of important covariates
```{r}
#age
table(donor_meta$age)
hist(donor_meta$age, breaks = seq(min(donor_meta$age),max(donor_meta$age),1),
     col="cornflowerblue", main = "Age Distribution across donors", xlab="age", ylab="number samples")

# sex
table(donor_meta$sex)
pie(table(donor_meta$sex), labels=paste(names(table(donor_meta$sex)), table(donor_meta$sex), sep=":"),main="Sex distribution across donors")

# tissues
table(donor_meta$n_tissues)
barplot(table(donor_meta$n_tissues), col="cornflowerblue", main="Number of tissues per donor")

# samples
table(donor_meta$n_samples)
barplot(table(donor_meta$n_samples), col="cornflowerblue", main="Number of samples per donor")

# tissues - detailed
df <- data.frame(table(rse_gene_ext$smtsd))
colnames(df) <- c("tissue", "number_samples")
df %<>% mutate(tissue=ifelse(tissue=="", "missing annotation", as.character(tissue)))
ggplot(filter(df, tissue !="missing annotation"), aes(x=tissue, y=number_samples, fill=tissue)) +geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=60, hjust=1)) +guides(fill=F)
df

 # tissues - broad
 df <- data.frame(table(rse_gene_ext$smts))
colnames(df) <- c("tissue", "number_samples")
df %<>% mutate(tissue=ifelse(tissue=="", "missing annotation", as.character(tissue)))
ggplot(filter(df, tissue !="missing annotation"), aes(x=tissue, y=number_samples, fill=tissue)) +geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=60, hjust=1)) +guides(fill=F)
 df
```


```{r}
sessionInfo()
```

